\documentclass[12pt, a4paper]{article}

% ── Packages ──────────────────────────────────────────────────────────────────
\usepackage[top=1in, bottom=1in, left=1.25in, right=1.25in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}
\usepackage{booktabs}
\usepackage{cite}
\usepackage{setspace}
\usepackage{microtype}   % tighter text, helps hit 5 pages precisely
\usepackage{parskip}     % spacing between paragraphs instead of indent

% ── Document metadata ─────────────────────────────────────────────────────────
\title{\textbf{An Introduction to Natural Language Processing}}
\author{[Your Name]}
\date{\today}

\onehalfspacing          % 1.5 line spacing

% ──────────────────────────────────────────────────────────────────────────────
\begin{document}

\maketitle
\thispagestyle{plain}

% ── 1. Introduction ───────────────────────────────────────────────────────────
\section{Introduction}

Natural Language Processing (NLP) is an interdisciplinary field situated at the
confluence of linguistics, computer science, and artificial intelligence. Its
central objective is to enable machines to read, interpret, and generate human
language in a manner that is both meaningful and contextually appropriate. Unlike
formal programming languages, natural language is inherently ambiguous, highly
context-dependent, and shaped by culture, intent, and pragmatic convention ---
properties that make its computational treatment a non-trivial scientific
challenge.

The practical significance of NLP is evident in the ubiquity of language-driven
technologies in everyday life. Modern search engines parse and rank billions of
documents against free-form queries; voice assistants such as Siri and Google
Assistant convert spoken utterances into actionable commands; neural machine
translation systems bridge hundreds of language pairs in near real time; and
conversational chatbots handle customer service interactions at scale. NLP thus
occupies a foundational position in the broader landscape of applied AI.

Historically, the field has passed through three broad paradigms. The earliest
systems, developed during the 1950s and 1960s, were \textit{rule-based}: linguists
manually encoded grammatical rules and lexical patterns into expert systems. The
1990s brought a \textit{statistical turn}, in which probabilistic models learned
regularities from large text corpora, reducing the reliance on hand-crafted
rules. The 2010s ushered in the \textit{deep learning era}, where neural networks
with distributed representations dramatically improved performance on virtually
every benchmark task. Since the early 2020s, the field has been dominated by
\textit{large language models} (LLMs) --- foundation models trained on
internet-scale data that exhibit remarkable few-shot generalisation
abilities~\cite{brown2020gpt3}.


% ── 2. Linguistic Foundations & Core Concepts ─────────────────────────────────
\section{Linguistic Foundations and Core Concepts}

A rigorous treatment of NLP requires familiarity with several levels of
linguistic analysis. \textbf{Morphology} concerns the internal structure of
words and the rules by which morphemes combine to form complex surface forms.
\textbf{Syntax} governs the arrangement of words into grammatically well-formed
phrases and sentences, typically captured through constituency or dependency
parse trees. \textbf{Semantics} addresses the literal meaning of linguistic
expressions, distinguishing word-level lexical semantics from sentence-level
compositional meaning. Finally, \textbf{pragmatics} considers how context,
speaker intent, and conversational conventions modulate the interpretation of an
utterance beyond its literal content.

Several core concepts recur throughout the NLP literature. A \textbf{token} is
the minimal unit of text fed to a model --- commonly a word, subword, or
character depending on the tokenisation scheme. A \textbf{corpus} is a
structured collection of text used for training or evaluation. An
\textbf{$n$-gram} is a contiguous sequence of $n$ tokens; unigrams, bigrams,
and trigrams are the most commonly used instances. The \textbf{vocabulary} $V$
is the set of all unique tokens present in a corpus; its cardinality directly
governs model dimensionality and computational cost.

Standardised benchmark datasets have played a central role in driving measurable
progress. The \textbf{Penn Treebank}~\cite{manning1999statistical} provides
syntactically annotated Wall Street Journal text widely used for parsing and
part-of-speech tagging research. \textbf{SQuAD} (Stanford Question Answering
Dataset) frames reading comprehension as span extraction over Wikipedia passages.
\textbf{GLUE} (General Language Understanding Evaluation) aggregates nine
diverse tasks --- including textual entailment, sentiment classification, and
coreference resolution --- into a single leaderboard, making it the de facto
standard for evaluating general-purpose language understanding models.


% ── 3. Text Preprocessing Pipeline ───────────────────────────────────────────
\section{Text Preprocessing Pipeline}

Raw text is seldom in a form suitable for direct modelling. A standard
preprocessing pipeline transforms unstructured strings into a clean, structured
representation through the following sequence of operations:

\begin{description}
  \item[Tokenisation.] The input string is segmented into individual tokens.
    Rule-based tokenisers split on whitespace and punctuation; subword methods
    such as Byte-Pair Encoding (BPE) further decompose rare words into
    frequent character sequences.
  \item[Lowercasing.] Converting all tokens to lower case reduces vocabulary
    size by collapsing case variants (e.g.\ \textit{NLP} and \textit{nlp}).
  \item[Stopword Removal.] High-frequency function words (e.g.\ \textit{the},
    \textit{is}, \textit{at}) are discarded when downstream tasks rely on
    content words alone.
  \item[Stemming vs.\ Lemmatisation.] Both steps reduce morphological variation.
    Stemming applies heuristic suffix-stripping (Porter Stemmer), whereas
    lemmatisation consults a morphological lexicon to return the canonical
    dictionary form.
  \item[Part-of-Speech (POS) Tagging.] Each token is labelled with its
    grammatical category (noun, verb, adjective, etc.), enabling syntax-aware
    downstream processing.
  \item[Named Entity Recognition (NER).] Spans of tokens denoting real-world
    entities --- persons, organisations, locations, dates --- are identified and
    classified, grounding text in factual knowledge.
\end{description}

Once preprocessed, text is converted into numerical representations. The
\textbf{Bag-of-Words (BoW)} model represents a document as a sparse vector of
raw term counts, discarding word order. \textbf{TF-IDF} reweights term counts
by inverse document frequency to down-weight terms that appear in many documents
and thus carry little discriminative information:

\[
  \text{TF-IDF}(t,d) = \text{TF}(t,d) \times \log\frac{N}{df(t)}
\]

\noindent where $\text{TF}(t,d)$ is the frequency of term $t$ in document $d$,
$N$ is the total number of documents in the corpus, and $df(t)$ is the number
of documents containing $t$. TF-IDF vectors remain a competitive baseline for
document retrieval and text classification tasks.


% ── 4. Classical NLP Approaches ───────────────────────────────────────────────
\section{Classical NLP Approaches}

The earliest NLP systems were built around hand-crafted linguistic rules.
Finite-state transducers, regular grammars, and context-free parsing algorithms
encoded expert knowledge about morphology and syntax into explicit symbolic
representations. While rule-based systems can achieve high precision within
narrow domains, they suffer from well-documented brittleness: any linguistic
phenomenon not anticipated by the rule designer leads to failure. Scaling such
systems to open-domain language is practically infeasible owing to the
combinatorial complexity of natural language variation.

The statistical paradigm addressed these shortcomings by estimating model
parameters from data rather than encoding them manually. \textbf{Naïve Bayes}
classifiers apply Bayes' theorem with a conditional independence assumption
over token features, yielding a computationally efficient and surprisingly
effective baseline for text categorisation tasks such as spam filtering and
sentiment analysis~\cite{manning1999statistical}. \textbf{Hidden Markov Models
(HMMs)} brought a sequence-modelling perspective to NLP: by treating the
observed token sequence as emissions from a latent Markov chain of linguistic
states, HMMs proved effective for POS tagging and speech recognition via the
Viterbi decoding algorithm.

Nevertheless, classical statistical models are constrained by the
\textit{curse of dimensionality}. Because tokens are represented as one-hot
vectors over the vocabulary, the resulting feature space is both
high-dimensional and extremely sparse. Co-occurrence statistics become
unreliable for infrequent tokens, and the discrete nature of word identities
prevents the model from generalising across semantically related forms. These
limitations motivated the move towards dense, continuous representations in the
neural era.


% ── 5. Neural NLP & Word Embeddings ───────────────────────────────────────────
\section{Neural NLP and Word Embeddings}

The introduction of dense word representations --- commonly called
\textit{word embeddings} --- marked a decisive break from sparse symbolic
features. \textbf{Word2Vec}~\cite{mikolov2013word2vec} learns low-dimensional
embedding vectors by training a shallow neural network to predict contextual
words. In the Skip-gram variant, the model is trained to maximise the log
probability of surrounding context words given a centre word. The objective
over a corpus of $T$ tokens with context window $c$ is:

\[
  J(\theta) = -\frac{1}{T} \sum_{t=1}^{T}
              \sum_{\substack{-c \leq j \leq c \\ j \neq 0}}
              \log P(w_{t+j} \mid w_t)
\]

The resulting vectors encode semantic regularities as linear geometric
relations (e.g.\ $\vec{\text{king}} - \vec{\text{man}} + \vec{\text{woman}}
\approx \vec{\text{queen}}$). \textbf{GloVe} (Global Vectors for Word
Representation) achieves similar results by factorising the global word
co-occurrence matrix, offering a principled probabilistic interpretation
alongside competitive downstream performance.

Recurrent architectures extended neural NLP to variable-length sequences.
\textbf{Recurrent Neural Networks (RNNs)} maintain a hidden state that is
updated at each time step, theoretically capturing long-range dependencies.
In practice, vanilla RNNs suffer from the \textit{vanishing gradient problem}:
gradients decay exponentially over long sequences, preventing the model from
retaining information over many time steps. \textbf{Long Short-Term Memory
(LSTM)} networks mitigate this via explicit memory cells and learnable gating
mechanisms that selectively read, write, and erase information.

A further leap came with the \textit{attention mechanism}, which allows a model
to compute a soft, weighted alignment between every pair of positions in a
sequence. Rather than compressing an entire input into a fixed-length vector,
attention enables the decoder at each output step to directly focus on the most
relevant input positions, yielding substantial gains in neural machine
translation and becoming the conceptual precursor to the Transformer.


% ── 6. Transformer Architecture & LLMs ───────────────────────────────────────
\section{Transformer Architecture and Large Language Models}

Vaswani et al.~\cite{vaswani2017attention} introduced the Transformer
architecture in 2017, dispensing entirely with recurrence in favour of
\textit{self-attention}. Given queries $Q$, keys $K$, and values $V$ derived
from linear projections of the input, scaled dot-product attention is computed
as:

\[
  \text{Attention}(Q,K,V) =
    \operatorname{softmax}\!\left(\frac{QK^{\top}}{\sqrt{d_k}}\right)V
\]

\noindent where $d_k$ is the dimensionality of the key vectors. Dividing by
$\sqrt{d_k}$ stabilises gradients by preventing the dot products from growing
large in magnitude. Multi-head attention runs $h$ parallel attention functions
and concatenates their outputs, allowing the model to jointly attend to
information from different representation subspaces.

The Transformer spawned two dominant architectural families.
\textbf{Encoder-only} models such as BERT~\cite{devlin2018bert} process the
full input bidirectionally and excel at understanding tasks --- classification,
NER, question answering. BERT is pretrained via masked language modelling, where
randomly masked tokens must be predicted from their context. \textbf{Decoder-only}
models such as GPT and its successors are autoregressive: they predict each
token conditioned on all preceding tokens, making them natural generative models
suitable for open-ended text generation.

Both families follow the \textit{pretraining--fine-tuning paradigm}: a model
is first trained on large amounts of unlabelled text, acquiring broad
linguistic knowledge, and then adapted to a specific task using a small labelled
dataset. More recently, \textit{prompt engineering} --- crafting natural language
instructions to elicit desired model behaviour without updating weights ---
and \textit{Retrieval-Augmented Generation (RAG)}, which conditions generation
on dynamically retrieved documents, have emerged as lightweight alternatives to
full fine-tuning, particularly relevant for knowledge-intensive tasks where
factual currency matters.


% ── 7. Key Tasks & Applications ───────────────────────────────────────────────
\section{Key Tasks and Applications}

NLP encompasses a broad spectrum of tasks, each with its own evaluation
protocols and state-of-the-art methods. Table~\ref{tab:tasks} summarises
representative tasks, the dominant techniques applied to them, and concrete
example applications.

\begin{table}[h]
\centering
\caption{Representative NLP Tasks, Techniques, and Applications}
\label{tab:tasks}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Task} & \textbf{Technique} & \textbf{Example Application} \\
\midrule
Sentiment Analysis     & Fine-tuned BERT / RNN       & Product review classification \\
Machine Translation    & Seq2Seq Transformer         & Google Translate              \\
Named Entity Recognition & CRF / Token Classification & Clinical record extraction   \\
Text Summarisation     & Encoder-decoder LLM         & News article summarisation    \\
Question Answering     & Span extraction (BERT)      & SQuAD, search engine snippets \\
Text Classification    & Naïve Bayes / GPT fine-tune & Spam filtering, topic labelling \\
\bottomrule
\end{tabular}
\end{table}

The diversity of these applications underscores the pervasive role of NLP
across industry verticals including healthcare, finance, legal technology, and
digital media. Notably, the same pretrained backbone model can be adapted to
many of these tasks with minimal task-specific data, making large pretrained
models extraordinarily cost-effective in practice.


% ── 8. Challenges & Ethics ────────────────────────────────────────────────────
\section{Challenges and Ethical Considerations}

Despite remarkable progress, NLP faces significant open challenges. Linguistic
\textit{ambiguity} --- lexical, syntactic, and referential --- remains difficult
to resolve without rich world knowledge and contextual grounding. The field is
also deeply skewed towards high-resource languages: fewer than 100 of the world's
approximately 7,000 languages have sufficient digital text to train competitive
models, raising concerns about linguistic exclusion.

Large language models are prone to \textit{hallucination} --- the confident
generation of factually incorrect or entirely fabricated statements --- a
phenomenon that limits their deployment in safety-critical domains. Ethical
concerns are equally pressing: models trained on web-scale corpora inherit
societal \textit{biases} related to gender, race, and religion, which can
propagate into downstream applications with harmful consequences. \textit{Privacy}
risks arise when models memorise personally identifiable information from training
data. Finally, the \textit{environmental cost} of pretraining frontier LLMs ---
measured in thousands of tonnes of CO$_2$ equivalent --- raises questions about
the long-term sustainability of the current scaling paradigm.


% ── 9. Conclusion ─────────────────────────────────────────────────────────────
\section{Conclusion}

Natural Language Processing has traversed a remarkable arc: from handcrafted
symbolic rules that could handle only narrow, pre-specified phenomena, through
probabilistic models that learned statistical regularities from data, to deep
neural architectures and, ultimately, to billion-parameter language models that
approach human-level performance on many benchmarks. The Transformer architecture
and the pretraining--fine-tuning paradigm have been the defining technical
contributions of the past decade, fundamentally reshaping what is possible at
the intersection of language and computation~\cite{vaswani2017attention,devlin2018bert}.

Looking ahead, the most consequential research directions include
\textit{multimodal NLP} --- integrating language with vision, audio, and
structured data; the development of \textit{efficient small models} that match
frontier performance with a fraction of the computational cost; and
\textit{alignment research}, which aims to ensure that language model behaviour
reliably reflects human values and intentions. As language increasingly mediates
the interface between humans and AI systems, NLP will remain one of the most
scientifically rich and socially consequential branches of artificial
intelligence research.


% ── References ────────────────────────────────────────────────────────────────
\begin{thebibliography}{9}

\bibitem{vaswani2017attention}
A.\ Vaswani, N.\ Shazeer, N.\ Parmar, J.\ Uszkoreit, L.\ Jones, A.\ N.\ Gomez,
\L.\ Kaiser, and I.\ Polosukhin,
``Attention is all you need,''
in \textit{Advances in Neural Information Processing Systems (NeurIPS)},
vol.\ 30, 2017.

\bibitem{devlin2018bert}
J.\ Devlin, M.-W.\ Chang, K.\ Lee, and K.\ Toutanova,
``BERT: Pre-training of deep bidirectional transformers for language
understanding,''
in \textit{Proceedings of NAACL-HLT}, pp.\ 4171--4186, 2019.

\bibitem{mikolov2013word2vec}
T.\ Mikolov, I.\ Sutskever, K.\ Chen, G.\ Corrado, and J.\ Dean,
``Distributed representations of words and phrases and their compositionality,''
in \textit{Advances in Neural Information Processing Systems (NeurIPS)},
vol.\ 26, 2013.

\bibitem{manning1999statistical}
C.\ D.\ Manning and H.\ Schütze,
\textit{Foundations of Statistical Natural Language Processing}.
Cambridge, MA: MIT Press, 1999.

\bibitem{brown2020gpt3}
T.\ B.\ Brown, B.\ Mann, N.\ Ryder, M.\ Subbiah, J.\ Kaplan, P.\ Dhariwal,
A.\ Neelakantan, \textit{et al.},
``Language models are few-shot learners,''
in \textit{Advances in Neural Information Processing Systems (NeurIPS)},
vol.\ 33, pp.\ 1877--1901, 2020.

\end{thebibliography}

\end{document}